---
title: "CytoGLMM extension" 
author: 
  name: "Anne-Maud Ferreira"
  affiliation: Department of Statistics, Stanford University
date: "`r Sys.Date()`"
package: CyTOFtoolbox
output: BiocStyle::html_document
vignette: >
    %\VignetteIndexEntry{CytoGLMM-extension}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}  
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=8, 
  fig.height=5
)
```

# Environment set up

```{r setup, warning = FALSE, message = FALSE}
library(CyTOFtoolbox)
library(CytoGLMM)
library(dplyr)
```

# Introduction

This is a short vignette showing an example of the volcano plot and modified
MDS plot.

# Data generation

We use the data generation process available in the `CytoGLMM` package:

```{r generate_data}
# Set seed
set.seed(23)
# Generate the data
df_raw <- generate_data()
# Extract protein names
protein_names <- names(df_raw)[3:12]
# Transform the data
df <- dplyr::mutate_at(df_raw, protein_names, function(x) asinh(x/5))
```

## MDS plot

The MDS plot provided by the `CytoGLMM` has been modified to add a threshold 
on how much the markers medians correlate with the MDS dimensions. The
input parameters `cor_threshold` is used to defined this threshold:

```{r mds_plot}
plot_MDS_withthreshold(df_samples = df,
                       protein_names = protein_names, 
                       sample_info_names = c("donor", "condition"), 
                       color = "condition",
                       cor_threshold = 0.2)
```

## Fit the GLMM model

Fit a Generalized Linear Mixed Model (GLMM) with donor random effects.

```{r fit_cytoglmm, warning = FALSE}
glmm_fit <- CytoGLMM::cytoglmm(df,
                              protein_names = protein_names,
                              condition = "condition",
                              group = "donor",
                              num_cores = 1)
```

Plot the output of the analysis:

```{r plot_fit_cytoglmm, warning = FALSE}
plot(glmm_fit, order = TRUE)
```

# Volcano plot

Plot the volcano plot as a *complementary* plot of the CytoGLMM model output.
As input parameters, we need:

- The raw data (i.e. untransformed data to compute the global Mean Signal Intensity (MSI));
- The output of the `CytoGLMM::cytoglmm` function. The output of the `CytoGLMM::cytoglm` 
model using the bootstrap can also be used as an input parameter;
- The condition tested in the `CytoGLMM` model.

Note that by default the p-value threshold equals $0.05$, it can be modify by 
using the input parameter alpha in the `prepare_data_for_volcanoplot` function.

```{r volcano_plot}
# Prepare the data
data_for_volplot <- prepare_data_for_volcanoplot(data = df_raw, 
                                                 protein_names = protein_names,
                                                 condition = "condition", 
                                                 CytoGLMM_fit = glmm_fit)
# Volcano plot
volcano_plot(data = data_for_volplot)
```

# Session information

```{r sessioninfo}
sessionInfo()
```