---
title: "Volcano plot"
author: "Anne-Maud Ferreira"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{volcano-plot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning = FALSE, message = FALSE}
library(CyTOFtoolbox)
library(CytoGLMM)
library(SummarizedExperiment)
library(magrittr)
library(tidyverse)
```

# Introduction

This is a short vignette showing an example of the volcano plot.

# Aghaeepour et al. (2017) dataset

As an example, the Aghaeepour et al. (2017) dataset was used:

- Selection of 16 women from the Aghaeepour et al. (2017) dataset 
- Samples from the first and third trimester of their pregnancy stimulated by IFN$\alpha$

```{r load_data}
# Path
zenodo_url <- "https://zenodo.org/record/2652578/files/"
cytof_data <- "se_aghaeepour2017immune.Rdata"
# Download data
## se_aghaeepour2017immune is an SummarizedExperiment object
load(url(paste0(zenodo_url, cytof_data)))
```

## Pre-processing of the data 

The data are processed as in the example of the CytoGLMM workflow available [online](https://christofseiler.github.io/CytoGLMM/articles/CytoGLMM.html#generalized-linear-model-with-bootstrap):


```{r process_data}
# Extract the values
exprs <- assay(se_aghaeepour2017immune)
# Extract the sample information
sample_info <- rowData(se_aghaeepour2017immune)
# Extract the sample names
sample_info_names <- names(sample_info)
# Create a data frame with the expression values and the sample information
df_samples <- cbind(as.data.frame(exprs), as.data.frame(sample_info))
df_samples %<>% as_tibble
# Extract protein names
protein_names <- colData(se_aghaeepour2017immune) %>% 
  as.data.frame %>% 
  dplyr::filter(type == "function") %>%
  .$protein_name

# Subset to NK cell population
df_samples_subset <- df_samples %>% dplyr::filter(celltype == "NK")
df_samples_subset %<>% dplyr::select(protein_names, sample_info_names)

# Transform the data
df_samples_subset_transformed <- df_samples_subset %>% 
  dplyr::mutate_at(protein_names, 
                   function(x) asinh(x/5))
```

## Fit the GLMM model

Fit a Generalized Linear Mixed Model (GLMM) with donor random effects.

```{r fit_cytoglmm, eval = FALSE}
# Does not run 
CytoGLMM_glmm_fit_aghaeepour2017immune <- 
  CytoGLMM::cytoglmm(df_samples_subset_transformed, 
                     protein_names = protein_names,
                     condition = "term", 
                     group = "donor",
                     cell_n_subsample = 1000, # Subsample is used ONLY for the sake of the data size
                     num_cores = 4)
```

The model was run outside of this vignette and store internally in the package, running exactly the previous chunks.

```{r load_fit_cytoglmm}
data("CytoGLMM_glmm_fit_aghaeepour2017immune")
```

# Volcano plot

Plot the volcano plot as a complementary plot for the CytoGLMM model output.
As input parameters, we need:

- The raw data (i.e. untransformed data to compute the global Mean Signal Intensity (MSI));
- The output of the `CytoGLMM::cytoglmm` function.

```{r volcano_plot}
# Prepare the data
data_for_volplot <- prepare_data_for_volcanoplot(data = df_samples_subset, 
                                                 protein_name = protein_names,
                                                 condition = "term", 
                                                 CytoGLMM_fit = CytoGLMM_glmm_fit_aghaeepour2017immune)
# Volcano plot
volcano_plot(data = data_for_volplot, 
             exp_conditions = c("1st trim", "3rd trim"))
```
